---
title: "Bee Wings Github"
author: "Jorge De La Cruz"
date: "2024-04-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

############################################################################################################

##Load Packages
```{r packages, warning=FALSE, echo=FALSE, message=FALSE}
library(htmltools)
library(rgl)
library(tidyverse)
library(geomorph) #NOTE: download XQuartz if using a Mac
library(devtools)
library(RRPP) #used for MANOVA 
library(adegenet) # used for DAPC
library(phangorn) #Maximum Likelihood (ML)
```

##Load TPS
```{r}
#Read in the TPS data for all bee wing images

all_tps_data <- readland.tps("data/bee_wings5.TPS",
                             specID = c("imageID"), readcurves = FALSE, 
                             warnmsg = TRUE,negNA = FALSE)

```

##Load csv Data
Read bee data by first reading all Hymenoptera. After this chunk of code, all bee data is saved as __df1__ and is ready to be merged with the TPS data.
```{r}
#need to have "beedata26jun22.csv" saved in working directory
bee_data <- read_csv("data/bee_metadata2.csv")

#clean up data. Only keep relevant columns
bee_data

#df1 <- bee_data
#df1 <- df1 %>% separate(catalogNumber, c("UCSB", "number", "side")) #issue btwn _ vs - in catalogNumber
#bee_data_clean <- df1 %>% tidyr::unite('catalogNumber', sep = "_", c('UCSB','number', 'side')) #solving it by switching to _ from - in #the downloaded Hymenoptera data

bee_data_clean <- bee_data
```

##Generalized Procrustes Analaysis (GPA)
Run GPA on TPS data, to then generate coordinate data for each landmark on each specimen. In this code, we start with tps data and end with a CSV of coordinate data for each specimen. This can then be merged with the bee data generated above.
```{r}
all_tps_gpa <- gpagen(all_tps_data, PrinAxes = TRUE)

write.csv(all_tps_gpa$coords, "data/tpsGPA.csv", row.names = TRUE)

```
###Adjust the Landmark Data to usable format
```{r}
lmdata <- read.csv("data/tpsGPA.csv") #read landmark data in

df_transpose = t(lmdata) #transpose the landmark data
df_transpose1 <- df_transpose
df_transpose1 <- df_transpose1[-c(1),] #Remove first row
#Format of LM data is currently two rows per specimen
#with one row being X coordinate values and one row
#being Y coordinate values. Below, we split into two 
#data frames and then stitch them back together to get
#18 different variables for the 9 landmarks
#(9 X coordinates and 9 Y coordinates)
lm1 <- df_transpose1
lmX <- lm1
lmY <- lm1
lmX1 <- lmX
lmX1 <- data.frame(lmX1)
lmX2 <- setNames(cbind(rownames(lmX1), lmX1, row.names = NULL),
                 c("name", "LM1x", "LM2x", "LM3x",
                   "LM4x", "LM5x", "LM6x", "LM7x", "LM8x", "LM9x"))
lmXonly <- lmX2[str_detect(lmX2$name, "X.UCSB"), ]
#repeat with Y
lmY1 <- lmY
lmY1 <- data.frame(lmY1)
lmY2 <- setNames(cbind(rownames(lmY1), lmY1, row.names = NULL),
                 c("name", "LM1y", "LM2y", "LM3y",
                   "LM4y", "LM5y", "LM6y", "LM7y", "LM8y", "LM9y"))
lmYonly <- lmY2[str_detect(lmY2$name, "Y.UCSB"), ]
lmXonly1 <- lmXonly
lmYonly1 <- lmYonly
lmXonly2 <- lmXonly1 %>% 
    tidyr::separate(name,                      
                    c("X","UCSB", "barcode", "wing","species", 
                      "location", "wingSide"), extra='drop') %>%
    tidyr::unite('catalogNumber', c('UCSB','barcode', 'wing')) 
lmXonly3 <- lmXonly2[,c(2,6:14)]

##now: lmXonly5 has x coordinate values for all 9 landmarks, and just the catalogNumber
###repeat for Y
lmYonly2 <- lmYonly1 %>% 
    tidyr::separate(name,                      
                    c("Y","UCSB", "barcode", "wing","species", 
                      "location", "wingSide"), extra='drop') %>%
    tidyr::unite('catalogNumber', c('UCSB','barcode', "wing")) 
lmYonly3 <- lmYonly2[,c(2,6:14)]

########
#Now: unite the Y and X coordinate dataframes into just one
lm_both <- merge(lmXonly3, lmYonly3, by=("catalogNumber"))
```

###Merge Landmark and Curation Data
```{r}
lm_data <- merge(lm_both, bee_data_clean, by=c("catalogNumber"), all = FALSE)
```

###Phylogeny Data
```{r}
#Mean LM values by genus to calculate phylogenetic signal

mean_lmx1 <- aggregate(LM1x ~ Genus, data = lm_data, FUN = mean)
mean_lmx2 <- aggregate(LM2x ~ Genus, data = lm_data, FUN = mean)
mean_lmx3 <- aggregate(LM3x ~ Genus, data = lm_data, FUN = mean)
mean_lmx4 <- aggregate(LM4x ~ Genus, data = lm_data, FUN = mean)
mean_lmx5 <- aggregate(LM5x ~ Genus, data = lm_data, FUN = mean)
mean_lmx6 <- aggregate(LM6x ~ Genus, data = lm_data, FUN = mean)
mean_lmx7 <- aggregate(LM7x ~ Genus, data = lm_data, FUN = mean)
mean_lmx8 <- aggregate(LM8x ~ Genus, data = lm_data, FUN = mean)
mean_lmx9 <- aggregate(LM9x ~ Genus, data = lm_data, FUN = mean)

mean_lmy1 <- aggregate(LM1y ~ Genus, data = lm_data, FUN = mean)
mean_lmy2 <- aggregate(LM2y ~ Genus, data = lm_data, FUN = mean)
mean_lmy3 <- aggregate(LM3y ~ Genus, data = lm_data, FUN = mean)
mean_lmy4 <- aggregate(LM4y ~ Genus, data = lm_data, FUN = mean)
mean_lmy5 <- aggregate(LM5y ~ Genus, data = lm_data, FUN = mean)
mean_lmy6 <- aggregate(LM6y ~ Genus, data = lm_data, FUN = mean)
mean_lmy7 <- aggregate(LM7y ~ Genus, data = lm_data, FUN = mean)
mean_lmy8 <- aggregate(LM8y ~ Genus, data = lm_data, FUN = mean)
mean_lmy9 <- aggregate(LM9y ~ Genus, data = lm_data, FUN = mean)

#mean_lmx3, mean_lmx4, mean_lmx5, mean_lmx6, mean_lmx7, mean_lmx8, mean_lmx9, mean_lmy1, mean_lmy2, mean_lmy3, mean_lmy4, mean_lmy5, mean_lmy6, mean_lmy7, mean_lmy8, mean_lmy9

mean_LM2 <- merge(mean_lmx1, mean_lmx2, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmx3, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmx4, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmx5, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmx6, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmx7, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmx8, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmx9, by=c("Genus"))

mean_LM2 <- merge(mean_LM2, mean_lmy1, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmy2, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmy3, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmy4, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmy5, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmy6, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmy7, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmy8, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmy9, by=c("Genus"))

rownames(mean_LM2) <- mean_LM2$Genus

mean_LM3 <- mean_LM2[, -1] 

write.csv(mean_LM3, file = "data/GM_Data.csv")
```

## MANOVA
Using MANOVA to test for difference between groups using the RRPP package.
```{r}
# MANOVA test with three species
manova_species <- lm_data
manova_species <- manova_species[,c(2:20)] #drop catalog number and other non LM variables
manova_species_data <- manova_species[,c(1:18)] #drop species variable
manova_species_data <- as.matrix(manova_species_data) #as matrix

fit_species <- lm.rrpp(manova_species_data ~ Species, SS.type = "I", 
               data = manova_species, print.progress = FALSE) #run the linear model fit
fitm_species <- manova.update(fit_species, print.progress = FALSE, tol = 0) #run manova update
summary(fitm_species, test = "Pillai") #summarize with a manova table
```

### DAPC (Discriminant Analysis of Principal Components)

## Cross-validation using DAPC
Assign "unknown" bees to the correct group, and test for % accuracy. Tells number of PCs to retain
```{r}
lm_values <- lm_data[,c(2:19)] # Only preserves LM values

xval <- xvalDapc(lm_values, lm_data$Subfamily, n.pca.max = 300, training.set = 0.9, result = "groupMean", center = TRUE, scale = FALSE, n.pca = NULL, n.rep = 30, xval.plot = TRUE)

xval$`Number of PCs Achieving Highest Mean Success` #Which number of PCs retained has highest accuracy?
xval$`Mean Successful Assignment by Number of PCs of PCA` #What is accuracy at various PCs retained?
```
#DAPC with 1 Discriminant Function
Rather than a PCA, we run a DAPC here to visualize difference between groups.
```{r}
dapc_1 <- dapc(lm_values,  
                grp = lm_data$Family, 
                n.pca = 10,
                n.da = 10) 

dapc_col <- c("red4", "darkblue", "forestgreen","darkcyan", "magenta4") #save color values for DAPC

scatter(dapc_1,1,1,bg="white", #1,1, = 1st discriminant function
        cstar=0, scree.pca=FALSE, posi.pca="topright",
        legend=TRUE, size.leg=4, posi.leg="topright", col = dapc_col) # plot w/ just one axis meaning just 1 DA

```
# DAPC with 2 Discriminant Functions
```{r}
population <- lm_data[,c(2:19)] # Only preserves LM values

dapc_2 <- dapc(population, 
                grp = lm_data$Family, #Species, Genus, Subfamily, Family, 
                n.pca = 11, #Axes retained in Principal Component Analysis (PCA)
                n.da = 10) #Axes retained in Discriminant Analysis

scatter(dapc_2, 1,2, #1,2, = 1st and 2nd discriminant functions
        scree.da = FALSE, posi.da = "topleft", 
        bg="white", 
        cstar=0,  pch=20, solid=.4, cex=2.5,scree.pca=FALSE, posi.pca="topleft",
        legend=TRUE, posi.leg = "bottomright",
        col = dapc_col) #plot the two populations

```

#Family DAPC
```{r}
population <- lm_data[,c(2:19)] # Only preserves LM values

excluded_families <- c() #Subset family "X" from data
family_data <- population[!lm_data$Family %in% excluded_families, ]
family_groups <- lm_data$Family[!lm_data$Family %in% excluded_families]

dapc_fam <- dapc(family_data, 
                grp = family_groups, 
                n.pca = 10, #Axes retained in Principal Component Analysis (PCA)
                n.da = 10) #Axes retained in Discriminant Analysis

phylo_col <- c("red4", "turquoise3", "gold2","forestgreen", "magenta4") #save colors for population

scatter(dapc_fam, scree.da = FALSE, posi.da = "bottomleft",
        bg="white", pch=20, 
        cstar=0, solid=.4, cex=2.5, scree.pca=FALSE, posi.pca="topbottom",
        legend=TRUE, posi.leg = "bottomright",
        col = dapc_col)


```

#Subfamily DAPC
```{r}
population <- lm_data[,c(2:19)] # Only preserves LM values

excluded_subfamilies <- c() #Subset subfamily "X" from data
filtered_subdata <- population[!lm_data$Subfamily %in% excluded_subfamilies, ]
filtered_subgroups <- lm_data$Subfamily[!lm_data$Subfamily %in% excluded_subfamilies]

dapc_sub <- dapc(filtered_subdata, 
                grp = filtered_subgroups, 
                n.pca = 12, #Axes retained in Principal Component Analysis (PCA)
                n.da = 10) #Axes retained in Discriminant Analysis

sub_col <- c("red4", "darkblue", "darkcyan", "forestgreen","chartreuse4", "gold3","maroon3", "magenta4")
                 

#("red4", "darkorange2","gold3", "seagreen3", "forestgreen","darkcyan","darkblue","magenta4")
scatter(dapc_sub, scree.da = FALSE, posi.da = "bottomleft",
        bg="white", pch=20, 
        cstar=0, solid=.45, cex=2.5, clab = 0.8, scree.pca=FALSE, posi.pca="topbottom",
        legend=TRUE, posi.leg = "bottomright",
        col = sub_col)


```

#Genus DAPC
```{r}
population <- lm_data[,c(2:19)] # Only preserves LM values

excluded_genus <- c() #Subset genus "X" from data
genus_data <- population[!lm_data$Family %in% excluded_genus, ]
genus_groups <- lm_data$Genus[!lm_data$Family %in% excluded_genus]

#included_genus <- c("Apis", "Xylocopa","Bombus", "Ceratina", "Eucera", "Diadasia", "Lasioglossum") #Subset "X" species from data
#genus_data <- population[lm_data$Genus %in% included_genus, ]
#genus_groups <- lm_data$Genus[lm_data$Genus %in% included_genus]


dapc_gen <- dapc(genus_data, 
                grp = genus_groups, 
                n.pca = 6, #Axes retained in Principal Component Analysis (PCA)
                n.da = 10) #Axes retained in Discriminant Analysis

genus_col <- c("red4", "darkblue", "darkcyan", "forestgreen","chartreuse4", "gold3","maroon3", "magenta4") #save colors for population


scatter(dapc_gen, scree.da = TRUE, posi.da = "bottomleft",
        bg="white", pch=20, 
        cstar=0, solid=.45, cex=2.5, clab = 0.4, scree.pca=FALSE, posi.pca="topbottom",
        legend=TRUE, cleg = 0.8, posi.leg = "bottomright",
        col = genus_col)


```

#Species DAPC
```{r}
population <- lm_data[,c(2:19)] # Only preserves LM values

#excluded_species <- c() #Subset "X" species from data
species_data <- population[!lm_data$Species %in% excluded_species, ]
species_groups <- lm_data$Species[!lm_data$Species %in% excluded_species]

#included_species <- c("Apis mellifera", "Xylocopa tabaniformis orpifex","Bombus californicus", "Bombus vosnesenskii", "Eucera actuosa") #Subset "X" species from data
#species_data <- population[lm_data$Species %in% included_species, ]
#species_groups <- lm_data$Species[lm_data$Species %in% included_species]

dapc_spe <- dapc(species_data, 
                grp = species_groups,
                n.pca = 11, #Axes retained in Principal Component Analysis (PCA)
                n.da = 10) #Axes retained in Discriminant Analysis

species_col <- c("red4", "darkblue", "darkcyan", "forestgreen","chartreuse4", "gold3","maroon3", "magenta4") #save colors for population

scatter(dapc_spe, scree.da = FALSE, posi.da = "bottomleft",
        bg="white", pch=20, 
        cstar=0, solid=.45, cex=2, clab = 0.5, scree.pca=FALSE, posi.pca="bottomleft",
        legend=TRUE, cleg = 0.6, posi.leg = "bottomright",
        col = species_col)



```

###3-Dimensional PCA - Family
```{r}
lm_data$Family <- factor(lm_data$Family,
                       levels = c("Apidae","Andrenidae","Megachilidae", "Halictidae"))

colors <- as.numeric(lm_data$Family)

head(lm_data)

pc <- princomp(lm_data[,2:19], cor=TRUE, scores=TRUE)

plot(pc,type="lines")

biplot(pc)

plot3d(pc$scores[,1:3], col = colors, size = 5, xlab = "PC1", ylab = "PC2", zlab = "PC3")

legend3d("topright", legend = levels(lm_data$Family), col = 1:length(levels(lm_data$Family)), pch = 16)

```
###3-Dimensional PCA - Subfamily
```{r}
lm_data$Subfamily <- factor(lm_data$Subfamily)

colors <- as.numeric(lm_data$Subfamily)

head(lm_data)

pc <- princomp(lm_data[,2:19], cor=TRUE, scores=TRUE)

plot(pc,type="lines")

biplot(pc)

plot3d(pc$scores[,1:3], col = colors, size = 5, xlab = "PC1", ylab = "PC2", zlab = "PC3")

legend3d("topright", legend = levels(lm_data$Subfamily), col = 1:length(levels(lm_data$Subfamily)), pch = 16)
```
###3-Dimensional PCA - Genus
```{r}
lm_data$Genus <- factor(lm_data$Genus)

colors <- as.numeric(lm_data$Genus)

head(lm_data)

pc <- princomp(lm_data[,2:19], cor=TRUE, scores=TRUE)

plot(pc,type="lines")

biplot(pc)

plot3d(pc$scores[,1:3], col = colors, size = 5, xlab = "PC1", ylab = "PC2", zlab = "PC3")

legend3d("topright", legend = levels(lm_data$Genus), col = 1:length(levels(lm_data$Genus)), pch = 16)
```



###Estimating phylogenetic tree from a character matrix by maximum-likelihood
```{r}
seq_data <- matrix(lm_data)

?phyDat
seq_phyDat <- phyDat(seq_data, type = "DNA")
dist_matrix <- dist.ml(seq_phyDat)

# Step 1: Create a starting tree using Neighbor-Joining (NJ)
nj_tree <- nj(dist_matrix)  # Using the distance matrix from earlier

# Step 2: Optimize the tree using Maximum Likelihood
ml_tree <- optim.pml(nj_tree, model = "GTR", rearrangement = "stochastic", control = pml.control(trace = 0))

# View the tree
plot(ml_tree$tree, main = "Maximum Likelihood Phylogeny")
```




