---
title: "Bee Tree"
author: "Jorge De La Cruz"
date: "2024-05-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Load Necessary Packages
```{r}
library(geomorph)
library(phytools)
library(ape)
library(ggtree)
library(vegan)
library(geiger)
```


```{r}
bee_tree <- ape::read.tree("data/BEE_mat7gen_p8pmAa_fst.nwk"); phytools::read.newick("data/BEE_mat7gen_p8pmAa_fst.nwk")

plot(bee_tree, cex = 0.1, no.margin=TRUE)

species_to_keep <- c("Apis_mellifera~Apidae~Apinae~Apini", "Bombus_impatiens~Apidae~Apinae~Bombini", "Agapostemon_tyleri~Halictidae~Halictinae~Caenohalictini", "Xylocopa_virginica~Apidae~Xylocopinae~Xylocopini", "Osmia_bucephala~Megachilidae~Megachilinae~Osmiini", "Megachile_rotundata~Megachilidae~Megachilinae~Megachilini", "Diadasia_chilensis~Apidae~Eucerinae~Emphorini", "Eucera_nigrilabris~Apidae~Eucerinae~Eucerini", "Andrena_nigroaenea~Andrenidae~Andreninae~Andrenini", "Halictus_ligatus~Halictidae~Halictinae~Halictini", "Ceratina_calcarata~Apidae~Xylocopinae~Ceratinini", "Habropoda_laboriosa~Apidae~Anthophorinae~Anthophorini", "Anthophora_montana~Apidae~Anthophorinae~Anthophorini", "Panurginus_albopilosus~Andrenidae~Panurginae~Panurgini", "Lasioglossum_albipes~Halictidae~Halictinae~Halictini")

pruned_tree <- drop.tip(bee_tree, setdiff(bee_tree$tip.label, species_to_keep)) # change "phylo_tree" to be the name of your tree

#closest_genus_Bombus <- which(pruned_tree$tip.label == "Bombus_impatiens~Apidae~Apinae~Bombini")
#new_bee_tree <- bind.tip(pruned_tree, "Bombus californicus", where = closest_genus_Bombus)

# Replace "current_name1" and "current_name2" with the current names of the species
# Replace "new_name1" and "new_name2" with the new names you want to assign
#tree$tip.label[tree$tip.label == "current_name2"] <- "new_name2"

pruned_tree$tip.label[pruned_tree$tip.label == "Apis_mellifera~Apidae~Apinae~Apini"] <- "Apis"
pruned_tree$tip.label[pruned_tree$tip.label == "Bombus_impatiens~Apidae~Apinae~Bombini"] <- "Bombus"
pruned_tree$tip.label[pruned_tree$tip.label == "Xylocopa_virginica~Apidae~Xylocopinae~Xylocopini"] <- "Xylocopa"
pruned_tree$tip.label[pruned_tree$tip.label == "Osmia_bucephala~Megachilidae~Megachilinae~Osmiini"] <- "Osmia"
pruned_tree$tip.label[pruned_tree$tip.label == "Megachile_rotundata~Megachilidae~Megachilinae~Megachilini"] <- "Megachile"
pruned_tree$tip.label[pruned_tree$tip.label == "Eucera_nigrilabris~Apidae~Eucerinae~Eucerini"] <- "Eucera"
pruned_tree$tip.label[pruned_tree$tip.label == "Diadasia_chilensis~Apidae~Eucerinae~Emphorini"] <- "Diadasia"
pruned_tree$tip.label[pruned_tree$tip.label == "Agapostemon_tyleri~Halictidae~Halictinae~Caenohalictini"] <- "Agapostemon"
pruned_tree$tip.label[pruned_tree$tip.label == "Andrena_nigroaenea~Andrenidae~Andreninae~Andrenini"] <- "Andrena"
pruned_tree$tip.label[pruned_tree$tip.label == "Halictus_ligatus~Halictidae~Halictinae~Halictini"] <- "Halictus"
pruned_tree$tip.label[pruned_tree$tip.label == "Ceratina_calcarata~Apidae~Xylocopinae~Ceratinini"] <- "Ceratina"
pruned_tree$tip.label[pruned_tree$tip.label == "Habropoda_laboriosa~Apidae~Anthophorinae~Anthophorini"] <- "Habropoda"
pruned_tree$tip.label[pruned_tree$tip.label == "Anthophora_montana~Apidae~Anthophorinae~Anthophorini"] <- "Anthophora"
pruned_tree$tip.label[pruned_tree$tip.label == "Panurginus_albopilosus~Andrenidae~Panurginae~Panurgini"] <- "Panurginus"
pruned_tree$tip.label[pruned_tree$tip.label == "Lasioglossum_albipes~Halictidae~Halictinae~Halictini"] <- "Lasioglossum" 




#ggtree(pruned_tree, cex = 0.1) + geom_tiplab() + ggplot2::xlim(0, 0.4)
ggtree(pruned_tree, cex = 0.5) + geom_tiplab(align = TRUE) + ggplot2::xlim(0, 0.3)

pruned_tree
```


```{r}
species_keep <- c("Apis_mellifera~Apidae~Apinae~Apini", "Bombus_impatiens~Apidae~Apinae~Bombini", "Agapostemon_tyleri~Halictidae~Halictinae~Caenohalictini", "Xylocopa_virginica~Apidae~Xylocopinae~Xylocopini", "Osmia_bucephala~Megachilidae~Megachilinae~Osmiini")

phylotree <- drop.tip(bee_tree, setdiff(bee_tree$tip.label, species_keep))

phylotree$tip.label[phylotree$tip.label == "Agapostemon_tyleri~Halictidae~Halictinae~Caenohalictini"] <- "Agapostemon"
phylotree$tip.label[phylotree$tip.label == "Apis_mellifera~Apidae~Apinae~Apini"] <- "Apis"
phylotree$tip.label[phylotree$tip.label == "Bombus_impatiens~Apidae~Apinae~Bombini"] <- "Bombus"
phylotree$tip.label[phylotree$tip.label == "Osmia_bucephala~Megachilidae~Megachilinae~Osmiini"] <- "Osmia"
phylotree$tip.label[phylotree$tip.label == "Xylocopa_virginica~Apidae~Xylocopinae~Xylocopini"] <- "Xylocopa"

#ggtree(pruned_tree, cex = 0.1) + geom_tiplab() + ggplot2::xlim(0, 0.4)
ggtree(phylotree, cex = 0.1, ladderize = TRUE) + geom_tiplab() + ggplot2::xlim(0, 0.2)

```


```{r}
phylo_distances <- cophenetic.phylo(pruned_tree)

phylo_distances
```
##Load TPS
```{#r}
#Read in the TPS data for all bee wing images

phylogeny_tps <- readland.tps("C:/Users/super/OneDrive/Desktop/Files/Beesearch/Files/tpsdig/Wing TPS/Phylogeny/PhylogenyTPS.TPS",
                             specID = c("imageID"), readcurves = FALSE, 
                             warnmsg = TRUE,negNA = FALSE)

phylogeny_tps

Agapostemon_tps <- readland.tps("C:/Users/super/OneDrive/Desktop/Files/Beesearch/Files/tpsdig/Wing TPS/Phylogeny/Agapostemon.TPS",
                             specID = c("imageID"), readcurves = FALSE, 
                             warnmsg = TRUE,negNA = FALSE)
Apis_tps <- readland.tps("C:/Users/super/OneDrive/Desktop/Files/Beesearch/Files/tpsdig/Wing TPS/Phylogeny/Apis.TPS",
                             specID = c("imageID"), readcurves = FALSE, 
                             warnmsg = TRUE,negNA = FALSE)
Bombus_tps <- readland.tps("C:/Users/super/OneDrive/Desktop/Files/Beesearch/Files/tpsdig/Wing TPS/Phylogeny/Bombus.TPS",
                             specID = c("imageID"), readcurves = FALSE, 
                             warnmsg = TRUE,negNA = FALSE)
Osmia_tps <- readland.tps("C:/Users/super/OneDrive/Desktop/Files/Beesearch/Files/tpsdig/Wing TPS/Phylogeny/Bombus.TPS",
                             specID = c("imageID"), readcurves = FALSE, 
                             warnmsg = TRUE,negNA = FALSE)
Xylocopa_tps <- readland.tps("C:/Users/super/OneDrive/Desktop/Files/Beesearch/Files/tpsdig/Wing TPS/Phylogeny/Xylocopa.TPS",
                             specID = c("imageID"), readcurves = FALSE, 
                             warnmsg = TRUE,negNA = FALSE)
```



##Generalized Procrustes Analaysis (GPA)
Run GPA on TPS data, to then generate coordinate data for each landmark on each specimen. In this code, we start with tps data and end with a CSV of coordinate data for each specimen. This can then be merged with the bee data generated above.
```{r}
phylogeny_gpa <- gpagen(phylogeny_tps, PrinAxes = TRUE)

# Assuming gmm_data is a 3D array (p x k x n): p = landmarks, k = dimensions, n = specimens
aligned_shapes <- phylogeny_gpa$coords

```
### 2. Prepare Your Data
Ensure you have:
-A phylogenetic tree (`phylo_tree`).
-A geometric morphometric dataset (`gmm_data`) with species names corresponding to the tips of the tree.


### 4. Calculate Morphometric Distances
Calculate pairwise Procrustes distances between species.

```{r}
# Calculate mean shapes for each species
mean_shapes <- mshape(aligned_shapes)
# Calculate Procrustes distances
procrustes_distances <- as.matrix(dist(mean_shapes))

mean_shapes
```


### 5. Calculate Phylogenetic Distances
Extract the cophenetic (phylogenetic) distances from the phylogenetic tree.
```{r}
phylo_distances <- cophenetic.phylo(phylotree)

phylo_distances
```


### 6. Mantel Test
Perform a Mantel test to assess the correlation between morphometric and phylogenetic distances.
```{r}
# Perform the Mantel test
procrustes_distances <- unlist(phylo_distances)



mantel_result <- mantel(phylo_distances, GM_Data)
# Print the results
print(mantel_result)
```


### 7. Phylogenetic Signal in Morphometric Data
Alternatively, you can test for phylogenetic signal in the 
morphometric data using methods like Blomberg’s K or Pagel’s λ.

```{r}
GM_Data <- read.csv("data/GM_Data.csv") 


phylotree
row.names(GM_Data)

row.names(GM_Data) <- GM_Data$X

# Remove species column 'X'
GM_Data$X <- NULL

GM_phylo <- as.matrix(GM_Data)



#Calculate Physignal (physignal - geomorph)
bee_signal <- physignal(GM_phylo, pruned_tree)

# Estimate effect size
effect_size <- physignal.z(A = GM_phylo, phy = pruned_tree)

# Print results
summary(bee_signal)
print(effect_size)

# Calculate Pagel’s λ (phylosig - phytools)
lambda_result <- phylosig(pruned_tree, GM_phylo, method = "lambda", test = TRUE)
print(lambda_result)

# Calculate Blomberg's K
K_result <- phylosig(pruned_tree, GM_phylo, method = "K", test = TRUE)
print(K_result)

# Calculate Pagel's λ (fitContinuous - geiger)
fitContinuous(pruned_tree, GM_phylo, model="lambda")

# Calculate Pagel's λ (fitContinuous - geiger)
fitContinuous(pruned_tree, GM_phylo, model="BM")


```

### Example Code
Here’s a more comprehensive example combining these steps:
```R
library(geomorph)
library(phytools)
library(ape)
# Load your data
phylo_tree <- read.tree("path_to_your_tree_file")
gmm_data <- read.csv("path_to_your_gmm_data_file")
# Perform GPA
gpa_result <- gpagen(gmm_data)
aligned_shapes <- gpa_result$coords
# Calculate mean shapes for each species
mean_shapes <- mshape(aligned_shapes)
# Calculate Procrustes distances
procrustes_distances <- as.matrix(dist(mean_shapes))
# Calculate phylogenetic distances
phylo_distances <- cophenetic.phylo(phylo_tree)
# Perform Mantel test
mantel_result <- mantel(procrustes_distances, phylo_distances)
print(mantel_result)
# Test for phylogenetic signal (Blomberg's K)
K_result <- phylosig(phylo_tree, mean_shapes, method = "K")
print(K_result)
# Test for phylogenetic signal (Pagel's lambda)
lambda_result <- phylosig(phylo_tree, mean_shapes, method = "lambda")
print(lambda_result)
```
### Notes:
- Ensure your species names match exactly between the morphometric dataset and the phylogenetic tree.
- The `phylomorphospace` function in `phytools` can visualize the phylogenetic relationships in morphospace.
This approach allows you to test the relationship between morphometric variation and phylogenetic distances, providing insight into how evolutionary history has influenced morphological traits.
