---
title: "Bee Tree - Github"
author: "Jorge De La Cruz"
date: "2025-08-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#install.packages("phytools")
#install.packages("ggtree")
#install.packages("geiger")
#install.packages("Quartet")
#install.packages("TreeDist")
#install.packages("Caper")


```


```{r}
library(phytools)
library(ape)
library(ggtree)
library(vegan)
library(geiger)
library(phangorn)
library(Quartet)
library(TreeDist)
#library(caper)

```

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("ggtree")
```



### 1. Load Necessary Packages
```{r}
library(geomorph)
library(phytools)
library(ape)
library(ggtree)
library(vegan)
library(geiger)
library(phangorn)
library(Quartet)
library(TreeDist)
library(caper)
```

###Load and Prune Phylogenetic Tree
```{r}
bee_tree <- ape::read.tree("data/BEE_mat7gen_p8pmAa_fst.nwk"); phytools::read.newick("data/BEE_mat7gen_p8pmAa_fst.nwk")

bee_infer <- ape::read.tree("data/BEE_mat7gen_p8pmAa_fst.nwk"); phytools::read.newick("data/BEE_mat7gen_p8pmAa_fst.nwk")

plot(bee_tree, cex = 0.1, no.margin=TRUE)

# Extract the genus from the tip labels
# Assuming the format is "genus_species~~family~~subfamily~tribe"
bee_tree$tip.label <- sub("_.*", "", bee_tree$tip.label)
bee_tree$tip.label <- sub("~.*", "", bee_tree$tip.label)
species_to_keep = c("Apis", "Bombus", "Agapostemon", "Xylocopa", "Osmia", "Megachile", "Diadasia", "Eucera", "Andrena", "Halictus", "Hesperapis", "Ceratina", "Habropoda", "Anthophora", "Panurginus", "Lasioglossum", "Calliopsis")

pruned_tree <- drop.tip(bee_tree, setdiff(bee_tree$tip.label, species_to_keep)) # change "phylo_tree" to be the name of your tree


#ggtree(pruned_tree, cex = 0.1) + geom_tiplab() + ggplot2::xlim(0, 0.4)
ggtree(pruned_tree, cex = 0.5) + geom_tiplab(align = TRUE) + ggplot2::xlim(0, 0.3)

ggtree(pruned_tree, branch.length="none", cex = 0.75) + geom_tiplab(size = 2) 

new_bee_tree <- pruned_tree

ggtree(new_bee_tree,branch.length="branch.length", cex = 0.75) + geom_tiplab(size = 2) 

write.tree(new_bee_tree, file = "output/newick.tre")
write.nexus(new_bee_tree, file = "data/bee_tree.nex")
```


```{r}
# 1. Read Data
mytree <- ape::read.tree("data/BEE_mat7gen_p8pmAa_fst.nwk"); phytools::read.newick("data/BEE_mat7gen_p8pmAa_fst.nwk") #species level tree
spec_jdlc <- #list of bee species of interest
  as.data.frame(c("Agapostemon texanus",
                             "Andrena oenotherae",
                             "Anthophora edwardsii",
                             "Apis mellifera",
                             "Bombus californicus",
                             "Bombus vosnesenskii",
                             "Calliopsis andreniformis",
                             #   "Ceratina punctagena",
                             "Ceratina strenua",
                             "Diadasia bituberculata",
                             "Eucera actuosa",
                             "Habropoda depressa",
                             "Halictus farinosus",
                             "Hesperapis larreae",
                             "Lasioglossum hitchensi",
                             "Megachile mendica",
                             "Megachile rotundata",
                             "Osmia nemoris",
                             #   "Panurginus polytrichus",
                             "Xylocopa tabaniformis"))

colnames(spec_jdlc)[1] <- "species" #update column name
mytree$tip.label <- sub("^([A-Za-z]+)_([a-z]+)_.*",
                        "\\1 \\2", mytree$tip.label) # Extract the species name from longer name 
species <- as.data.frame(mytree$tip.label)
tip.labels <- as.data.frame(mytree$tip.label)
colnames(species)[1] <- "species" #update column name


# 2. Prune Tree
tree_tips <- mytree$tip.label
bee_species <- c(spec_jdlc$species)
matching_names <- tree_tips[!tree_tips %in% bee_species]
trimmed_tree <- drop.tip(mytree, matching_names)
trimmed_tree$tip.label #check labels
plot(trimmed_tree) # should only include your species of interest
trimmed_tree_rt <- root(trimmed_tree, outgroup = "Hesperapis larreae", resolve.root = TRUE)
plot(trimmed_tree_rt)
branch_lengths <- trimmed_tree_rt$edge.length

plot.phylo(trimmed_tree_rt, cex = .7 main = "test plot")


# 3. Make Ultrametric (idk if you want to do this for your project or not??)
tree_ultra <- chronos(trimmed_tree) 
```



###Calculate Phylogenetic Distances
Extract the cophenetic (phylogenetic) distances from the phylogenetic tree.
Calculate pairwise Procrustes distances between species.
```{r}

phylo_distances <- cophenetic.phylo(pruned_tree)

phylo_distances

write.csv(phylo_distances, "data/phylo_distances.csv", row.names = TRUE)

```



### Phylogenetic Signal in Morphometric Data
Test for phylogenetic signal in the 
morphometric data using methods like Blomberg’s K or Pagel’s λ.

```{r}
CV_Data <- read.csv("data/CV_Data.csv") 

row.names(CV_Data)

row.names(CV_Data) <- CV_Data$Genus

# Remove species column 'X'
CV_Data$Genus <- NULL

CV_phylo <- as.matrix(CV_Data)

# Calculate Pagel’s λ (phylosig - phytools)
lambda_result <- phylosig(pruned_tree, CV_phylo, method = "lambda", test = TRUE)
print(lambda_result)

# Calculate Blomberg's K
K_result <- phylosig(pruned_tree, CV_phylo, method = "K", test = TRUE)
print(K_result)
```

#Phylogenetic Signal (SUBSET)
```{r}
GM_Data_subset <- read.csv("data/GM_Data_subset.csv") 

row.names(GM_Data_subset)

row.names(GM_Data_subset) <- GM_Data_subset$X

# Remove species column 'X'
GM_Data_subset$X <- NULL

GM_phylo_subset <- as.matrix(GM_Data_subset)

#Calculate Physignal (physignal - geomorph)
bee_signal <- physignal(GM_phylo_subset, pruned_tree)

# Estimate effect size
effect_size <- physignal.z(A = GM_phylo_subset, phy = pruned_tree)

# Print results
summary(bee_signal)
print(effect_size)

# Calculate Pagel’s λ (phylosig - phytools)
lambda_result <- phylosig(pruned_tree, GM_phylo_subset, method = "lambda", test = TRUE)
print(lambda_result)

# Calculate Blomberg's K
K_result <- phylosig(pruned_tree, GM_phylo_subset, method = "K", test = TRUE)
print(K_result)

# Calculate Pagel's λ (fitContinuous - geiger)
#fitContinuous(pruned_tree, GM_phylo, model="lambda")

# Calculate Pagel's λ (fitContinuous - geiger)
#fitContinuous(pruned_tree, GM_phylo, model="BM")

```


### Compare Phylogenetic Trees
```{r}
### Read trees

# Read known/reference tree
tree_known <- ape::read.nexus("data/bee_tree.nex")
ggtree(tree_known,branch.length="branch.length", cex = 0.75) + geom_tiplab(size = 2) 


# Read inferred tree

#Neighbor-Joining Analysis
tree_mesquite <- ape::read.nexus("data/tree_mesquite.nex")
ggtree(tree_mesquite,branch.length="branch.length", cex = 0.75) + geom_tiplab(size = 2) 

#UPGMA 
tree_UPGMA <- ape::read.nexus("data/tree_UPGMA.nex")
ggtree(tree_UPGMA,branch.length="branch.length", cex = 0.75) + geom_tiplab(size = 2) 


#Neighbor-Joining Analysis (SUBSET)
tree_NJ_subset <- ape::read.nexus("data/NJ_subset.nex")
ggtree(tree_NJ_subset,branch.length="branch.length", cex = 0.75) + geom_tiplab(size = 2) 


#UPGMA (SUBSET)
tree_UPGMA_subset <- ape::read.nexus("data/UPGMA_subset.nex")
ggtree(tree_UPGMA_subset,branch.length="branch.length", cex = 0.75) + geom_tiplab(size = 2) 


#####################


# Match taxa order for plotting
?cophylo
trees <- cophylo(tree_known, tree_mesquite, rotate = TRUE)
trees2 <- cophylo(tree_known, tree_UPGMA, rotate = TRUE)
trees3 <- cophylo(tree_mesquite, tree_UPGMA, rotate = TRUE)
trees4 <- cophylo(tree_known, tree_UPGMA_subset, rotate = TRUE)


# Plot
phylocompare <- plot(trees, link.type = "curved", link.lwd = 2, link.col = make.transparent("red", 0.5)) #NJ vs Reference tree 
plot(trees2, link.type = "curved", link.lwd = 2, link.col = make.transparent("red", 0.5)) #UPGMA vs Reference tree
plot(trees3, link.type = "curved", link.lwd = 2, link.col = make.transparent("red", 0.5)) #NJ vs UPGMA
plot(trees4, link.type = "curved", link.lwd = 2, link.col = make.transparent("red", 0.5)) #UPGMA (subset) vs Reference tree

#Save
# Open the SVG graphics device
svg("plots/phylo_comparison.svg", width = 12, height = 6) # You can adjust width and height

# Plot your phylogenetic tree
plot(trees, link.type = "curved", link.lwd = 3, link.col = make.transparent("red", 0.5)) # Replace with your actual plotting code

# Close the graphics device to save the file
dev.off()


```



#Calculate Quartet Tree Similarity
```{r}
#Quartet

#Reference Tree vs Neighbor-Joining Tree
RefvsNJ <- QuartetStatus(tree_known, tree_mesquite)

RefvsNJ

SimilarityMetrics(RefvsNJ, similarity = TRUE)

VisualizeQuartets(tree_known, tree_mesquite)


#Neighbor-Joining Tree vs UPGMA Tree
NJvsUPGMA <- QuartetStatus(tree_mesquite, tree_UPGMA)

SimilarityMetrics(NJvsUPGMA, similarity = TRUE)


#Reference Tree vs UPGMA Tree
RefvsUPGMA <- QuartetStatus(tree_known, tree_UPGMA)

SimilarityMetrics(RefvsUPGMA, similarity = TRUE)

VisualizeQuartets(tree_known, tree_UPGMA)



#Tree Info
tree_known
  
tree_mesquite

tree_UPGMA

```




```{r}
#Reference Tree vs Neighbor-Joining Tree (SUBSET)
RefvsNJsb <- QuartetStatus(tree_known, tree_NJ_subset)
SimilarityMetrics(RefvsNJsb, similarity = TRUE)

#Reference Tree vs UPGMA Tree (SUBSET)
RefvsUPGMAsb <- QuartetStatus(tree_known, tree_UPGMA_subset)

SimilarityMetrics(RefvsUPGMAsb, similarity = TRUE)

```









```{r}
#Neighbor-Joining Analysis (SUBSET) ROOTED
#tree_NJ_subset <- ape::read.nexus("data/NJ_subset_rooted.nex")
#plot(tree_NJ_subset)

#Neighbor-Joining Analysis (SUBSET)
tree_NJ_subset <- ape::read.nexus("data/NJ_subset.nex")
ggtree(tree_NJ_subset,branch.length="branch.length", cex = 0.75) + geom_tiplab(size = 2) 


#UPGMA (SUBSET)
tree_UPGMA_subset <- ape::read.nexus("data/UPGMA_subset.nex")
ggtree(tree_UPGMA_subset,branch.length="branch.length", cex = 0.75) + geom_tiplab(size = 2) 


#tree_NJ_subset_rooted <- ape::read.tree("data/NJ_subset_rooted2.nwk"); phytools::read.newick("data/NJ_subset_rooted2.nwk")

rooted_tree_outgroup <- root(tree_UPGMA_subset, outgroup = c("Hesperapis"), resolve.root = TRUE)


# Plot the rooted tree
plot(rooted_tree_outgroup)

ggtree(rooted_tree_outgroup,branch.length="branch.length", cex = 0.75) + geom_tiplab(size = 2) 

tree_NJ_subset_rooted



#Reference Tree vs Neighbor-Joining Tree
RefvsRoot <- QuartetStatus(tree_known, rooted_tree_outgroup)

SimilarityMetrics(RefvsRoot, similarity = TRUE)

```

#Phylogenetic Generalized Least Squares (PGLS)
```{r}
data(shorebird)
shorebird <- comparative.data(shorebird.tree, shorebird.data, Species, vcv=TRUE, vcv.dim=3)
mod1 <- pgls(log(Egg.Mass) ~ log(M.Mass) * log(F.Mass), shorebird, lambda='ML')
mod2 <- pgls(log(Egg.Mass) ~ log(M.Mass), data=shorebird, lambda='ML', delta='ML')

summary(mod1)
```

