---
title: "Bee R"
author: "Jorge De La Cruz"
date: "2025-11-24"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

############################################################################################################

##Load Packages
```{r packages, warning=FALSE, echo=FALSE, message=FALSE}
library(tidyverse)
library(htmltools)
library(rgl)
library(geomorph) #NOTE: download XQuartz if using a Mac
library(devtools)
library(RRPP) #used for MANOVA 
library(adegenet) # used for DAPC
library(phangorn) #Maximum Likelihood (ML)

library(phytools) #
library(ape) #
library(ggtree) #Display phylogenetic trees
library(vegan) #
library(geiger) #
library(Quartet) #Calculate quartet similarity of phylogenetic trees
library(TreeDist) #
```

##Load TPS
```{r}
#Read in the TPS data for all bee wing images

all_tps_data <- readland.tps("data/bee_wings6.TPS",
                             specID = c("imageID"), readcurves = FALSE, 
                             warnmsg = TRUE,negNA = FALSE)

```

##Load csv Data
Read bee data by first reading all Hymenoptera. After this chunk of code, all bee data is saved as __df1__ and is ready to be merged with the TPS data.
```{r}
#need to have "beedata26jun22.csv" saved in working directory
bee_data <- read_csv("data/bee_metadata2.csv")

#df1 <- bee_data
#df1 <- df1 %>% separate(catalogNumber, c("UCSB", "number", "side")) #issue btwn _ vs - in catalogNumber
#bee_data_clean <- df1 %>% tidyr::unite('catalogNumber', sep = "_", c('UCSB','number', 'side')) #solving it by switching to _ from - in #the downloaded Hymenoptera data

bee_data_clean <- bee_data
```

##Generalized Procrustes Analaysis (GPA)
Run GPA on TPS data, to then generate coordinate data for each landmark on each specimen. In this code, we start with tps data and end with a CSV of coordinate data for each specimen. This can then be merged with the bee data generated above.
```{r}
all_tps_gpa <- gpagen(all_tps_data, PrinAxes = TRUE)

write.csv(all_tps_gpa$coords, "data/tpsGPA.csv", row.names = TRUE)

```
###Adjust the Landmark Data to usable format
```{r}
lmdata <- read.csv("data/tpsGPA.csv") #read landmark data in

df_transpose = t(lmdata) #transpose the landmark data
df_transpose1 <- df_transpose
df_transpose1 <- df_transpose1[-c(1),] #Remove first row
#Format of LM data is currently two rows per specimen
#with one row being X coordinate values and one row
#being Y coordinate values. Below, we split into two 
#data frames and then stitch them back together to get
#18 different variables for the 9 landmarks
#(9 X coordinates and 9 Y coordinates)
lm1 <- df_transpose1
lmX <- lm1
lmY <- lm1
lmX1 <- lmX
lmX1 <- data.frame(lmX1)
lmX2 <- setNames(cbind(rownames(lmX1), lmX1, row.names = NULL),
                 c("name", "LM1x", "LM2x", "LM3x",
                   "LM4x", "LM5x", "LM6x", "LM7x", "LM8x", "LM9x"))
lmXonly <- lmX2[str_detect(lmX2$name, "X."), ]
#repeat with Y
lmY1 <- lmY
lmY1 <- data.frame(lmY1)
lmY2 <- setNames(cbind(rownames(lmY1), lmY1, row.names = NULL),
                 c("name", "LM1y", "LM2y", "LM3y",
                   "LM4y", "LM5y", "LM6y", "LM7y", "LM8y", "LM9y"))
lmYonly <- lmY2[str_detect(lmY2$name, "Y."), ]
lmXonly1 <- lmXonly
lmYonly1 <- lmYonly
lmXonly2 <- lmXonly1 %>% 
    tidyr::separate(name,                      
                    c("X","collection", "barcode", "wing","species", 
                      "location", "wingSide"), extra='drop') %>%
    tidyr::unite('catalogNumber', c('collection', 'barcode', 'wing')) 
lmXonly3 <- lmXonly2[,c(2,6:14)]

##now: lmXonly5 has x coordinate values for all 9 landmarks, and just the catalogNumber
###repeat for Y
lmYonly2 <- lmYonly1 %>% 
    tidyr::separate(name,                      
                    c("Y","collection", "barcode", "wing","species", 
                      "location", "wingSide"), extra='drop') %>%
    tidyr::unite('catalogNumber', c('collection', 'barcode', "wing")) 
lmYonly3 <- lmYonly2[,c(2,6:14)]

########
#Now: unite the Y and X coordinate dataframes into just one
lm_both <- merge(lmXonly3, lmYonly3, by=("catalogNumber"))
```


###Merge Landmark and Curation Data
```{r}
lm_data <- merge(lm_both, bee_data_clean, by=c("catalogNumber"), all = FALSE)

write.csv(lm_data, file = "data/LM_Data.csv")
```

## Create subset with equal numbers of specimens
Since there are different numbers of bee wings from each genus, we randomly select the same number of bees from each genus to create a subset with equal number of wings.
```{r}
subset_lm <- lm_data %>%
  group_by(Genus) %>%
  slice_sample(n = 10) %>%
  ungroup()

write.csv(subset_lm, file = "data/subset_data.csv")
```


## MANOVA
Using MANOVA to test for difference between groups using the RRPP package.
```{r}
# MANOVA test between genera
manova_genus <- subset_lm #subset
manova_genus <- manova_genus[,c(2:23)] #drop catalog number and other non LM variables
manova_genus_data <- manova_genus[,c(1:18)] #drop genus variable
manova_genus_data <- as.matrix(manova_genus_data) #as matrix

fit_genus <- lm.rrpp(manova_genus_data ~ Genus, SS.type = "I", #Species, Genus, Subfamily, Family
               data = manova_genus, print.progress = FALSE) #run the linear model fit
fitm_genus <- manova.update(fit_genus, print.progress = FALSE, tol = 0) #run manova update

#MANOVA Results
summary(fitm_genus, test = "Pillai") #summarize with a manova table
summary(fitm_genus, test = "Wilks") #summarize with a manova table
summary(fitm_genus, test = "Hotelling-Lawley") #summarize with a manova table

```


###Calculate mean LM values
```{r}
### SUBSET
#Mean LM values by genus to calculate phylogenetic signal

mean_lmx1 <- aggregate(LM1x ~ Genus, data = subset_lm, FUN = mean)
mean_lmx2 <- aggregate(LM2x ~ Genus, data = subset_lm, FUN = mean)
mean_lmx3 <- aggregate(LM3x ~ Genus, data = subset_lm, FUN = mean)
mean_lmx4 <- aggregate(LM4x ~ Genus, data = subset_lm, FUN = mean)
mean_lmx5 <- aggregate(LM5x ~ Genus, data = subset_lm, FUN = mean)
mean_lmx6 <- aggregate(LM6x ~ Genus, data = subset_lm, FUN = mean)
mean_lmx7 <- aggregate(LM7x ~ Genus, data = subset_lm, FUN = mean)
mean_lmx8 <- aggregate(LM8x ~ Genus, data = subset_lm, FUN = mean)
mean_lmx9 <- aggregate(LM9x ~ Genus, data = subset_lm, FUN = mean)

mean_lmy1 <- aggregate(LM1y ~ Genus, data = subset_lm, FUN = mean)
mean_lmy2 <- aggregate(LM2y ~ Genus, data = subset_lm, FUN = mean)
mean_lmy3 <- aggregate(LM3y ~ Genus, data = subset_lm, FUN = mean)
mean_lmy4 <- aggregate(LM4y ~ Genus, data = subset_lm, FUN = mean)
mean_lmy5 <- aggregate(LM5y ~ Genus, data = subset_lm, FUN = mean)
mean_lmy6 <- aggregate(LM6y ~ Genus, data = subset_lm, FUN = mean)
mean_lmy7 <- aggregate(LM7y ~ Genus, data = subset_lm, FUN = mean)
mean_lmy8 <- aggregate(LM8y ~ Genus, data = subset_lm, FUN = mean)
mean_lmy9 <- aggregate(LM9y ~ Genus, data = subset_lm, FUN = mean)

#mean_lmx3, mean_lmx4, mean_lmx5, mean_lmx6, mean_lmx7, mean_lmx8, mean_lmx9, mean_lmy1, mean_lmy2, mean_lmy3, mean_lmy4, mean_lmy5, mean_lmy6, mean_lmy7, mean_lmy8, mean_lmy9

mean_LM2 <- merge(mean_lmx1, mean_lmx2, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmx3, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmx4, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmx5, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmx6, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmx7, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmx8, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmx9, by=c("Genus"))

mean_LM2 <- merge(mean_LM2, mean_lmy1, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmy2, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmy3, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmy4, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmy5, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmy6, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmy7, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmy8, by=c("Genus"))
mean_LM2 <- merge(mean_LM2, mean_lmy9, by=c("Genus"))

rownames(mean_LM2) <- mean_LM2$Genus

mean_LM3 <- mean_LM2[, -1] 

LM_matrix_subset <- mean_LM3

write.csv(LM_matrix_subset, file = "data/GM_Data_subset.csv")

```


### DAPC (Discriminant Analysis of Principal Components)

## Cross-validation using DAPC
Assign "unknown" bees to the correct group, and test for % accuracy. Tells number of PCs to retain
```{r}
lm_values <- lm_data[,c(2:19)] # Only preserves LM values
lm_values_sub <- subset_lm[,c(2:19)] # Subset

#Replace lm_data$____ with taxonomic level 
xval <- xvalDapc(lm_values, lm_data$Genus, n.pca.max = 300, training.set = 0.9, result = "groupMean", center = TRUE, scale = FALSE, n.pca = NULL, n.rep = 30, xval.plot = TRUE)

#Replace lm_data$____ with taxonomic level (Subset)
xval2 <- xvalDapc(lm_values_sub, subset_lm$Genus, n.pca.max = 300, training.set = 0.9, result = "groupMean", center = TRUE, scale = FALSE, n.pca = NULL, n.rep = 30, xval.plot = TRUE)

xval$`Number of PCs Achieving Highest Mean Success` #Which number of PCs retained has highest accuracy? Put this value for n.pca = ___ when doing dapc() 

xval$`Mean Successful Assignment by Number of PCs of PCA` #What is accuracy at various PCs retained?

```

#DAPC with 1 Discriminant Function
Rather than a PCA, we run a DAPC here to visualize difference between groups.
```{r}
dapc_1 <- dapc(lm_values,  
                grp = lm_data$Family, 
                n.pca = 10,
                n.da = 10) 

dapc_col <- c("red4", "darkblue", "forestgreen","darkcyan", "magenta4") #save color values for DAPC

scatter(dapc_1,1,1,bg="white", #1,1, = 1st discriminant function
        cstar=0, scree.pca=FALSE, posi.pca="topright",
        legend=TRUE, size.leg=4, posi.leg="topright", col = dapc_col) # plot w/ just one axis meaning just 1 DA

```


# DAPC with 2 Discriminant Functions (SUBSET)
```{r}
population <- subset_lm[,c(2:19)] # Only preserves LM values

dapc_2 <- dapc(population, 
                grp = subset_lm$Genus, #Species, Genus, Subfamily, Family, 
                n.pca = 11, #Axes retained in Principal Component Analysis (PCA)
                n.da = 10) #Axes retained in Discriminant Analysis

scatter(dapc_2, 1, 2,  # 1st and 2nd discriminant functions
        scree.da = FALSE, posi.da = "topleft", 
        bg = "white", 
        cstar = 0, pch = 20, solid = 0.4, cex = 2.5, 
        scree.pca = FALSE, posi.pca = "topleft",
        legend = TRUE, posi.leg = "bottomright",
        col = dapc_col,
        clabel = 0.5)       # size of group labels

prcomp(population)

print.dapc(dapc_2)

summary.dapc(dapc_2)

predict.dapc(dapc_2)

```


### Bee Phylogeny

###Load and Prune Phylogenetic Tree
```{r}
bee_tree <- ape::read.tree("data/BEE_mat7gen_p8pmAa_fst.nwk"); phytools::read.newick("data/BEE_mat7gen_p8pmAa_fst.nwk")

# Extract the genus from the tip labels
# Assuming the format is "genus_species~~family~~subfamily~tribe"
bee_tree$tip.label <- sub("_.*", "", bee_tree$tip.label)
bee_tree$tip.label <- sub("~.*", "", bee_tree$tip.label)
species_to_keep = c("Apis", "Bombus", "Agapostemon", "Xylocopa", "Osmia", "Megachile", "Diadasia", "Eucera", "Andrena", "Halictus", "Hesperapis", "Ceratina", "Habropoda", "Anthophora", "Panurginus", "Lasioglossum", "Calliopsis")

pruned_tree <- drop.tip(bee_tree, setdiff(bee_tree$tip.label, species_to_keep)) 
ref_tree <- pruned_tree

write.tree(ref_tree, file = "output/newick.tre")
write.nexus(ref_tree, file = "data/bee_tree.nex")
```

###Calculate Phylogenetic Distances
Extract the cophenetic (phylogenetic) distances from the phylogenetic tree.
Calculate pairwise Procrustes distances between species.
```{r}
phylo_distances <- cophenetic.phylo(ref_tree)

write.csv(phylo_distances, "data/phylo_distances.csv", row.names = TRUE)
```


### Phylogenetic Signal in Morphometric Data (DELETE)
Test for phylogenetic signal in the 
morphometric data using methods like Blomberg’s K or Pagel’s λ.
```{r}
CV_Data <- read.csv("data/CV_Data.csv") 
row.names(CV_Data)
row.names(CV_Data) <- CV_Data$Genus
CV_Data$Genus <- NULL # Remove species column 'X'

CV_phylo <- as.matrix(CV_Data)

# Calculate Pagel’s λ (phylosig - phytools)
lambda_result <- phylosig(pruned_tree, CV_phylo, method = "lambda", test = TRUE)
print(lambda_result)

# Calculate Blomberg's K
K_result <- phylosig(pruned_tree, CV_phylo, method = "K", test = TRUE)
print(K_result)
```

###Canonical Variate Analysis
```{r}
library(Morpho) #package for geometric morphometrics and canonical variate analysis
library(car) #required to plot cva
library(shapes) #required to plot cva


lm_data <- read.csv("data/LM_Data.csv")
lm_data <- lm_data[,-1]

vari <- lm_data[,2:19]
facto <- as.factor(lm_data$Genus)
catalog <- as.factor(lm_data$catalogNumber)

cva.1 <- CVA(vari, groups=facto) #canonical variate analysis


## get the typicality probabilities and resulting classifications - tagging
## all specimens with a probability of < 0.01 as outliers (assigned to no class)
typprobs <- typprobClass(cva.1$CVscores,groups=facto)
print(typprobs)
```

#Visualize Canonical Variate Analysis
```{r}
grp <- levels(facto)
cols <- seq_along(grp)
pchs <- seq_along(grp)

plot(cva.1$CVscores,
     col = as.numeric(facto),#color
     pch = as.numeric(facto), #symbol
     type = "n",
     asp = 1,
     xlab = paste("1st canonical axis", round(cva.1$Var[1,2],1), "%"),
     ylab = paste("2nd canonical axis", round(cva.1$Var[2,2],1), "%"))

points(cva.1$CVscores,
       col = as.numeric(facto),#color
       pch = as.numeric(facto),#symbols
       cex = 0.6) #point size

legend("topright",
       inset = c(0, 0),
       legend = grp,
       col = cols,
       pch = pchs,
       cex = 0.75,
       bty = "n")

#text(cva.1$CVscores, #Label each data point
#     labels = as.character(catalog), #catalog number
#     col = as.numeric(facto),
#     cex = 0.2)
```

### Phylogenetic Signal in Morphometric Data
Test for phylogenetic signal in the morphometric data using methods like Blomberg’s K or Pagel’s λ.
```{r}
library(phytools)

LMmeans <- cva.1$groupmeans

# Calculate Pagel’s λ (phylosig - phytools)
lambda_result <- phylosig(ref_tree, LMmeans, method = "lambda", test = TRUE)
print(lambda_result)

# Calculate Blomberg's K
K_result <- phylosig(ref_tree, LMmeans, method = "K", test = TRUE)
print(K_result)
```

###Morphometric Phylogenetic Tree Reconstruction
```{r}
library(phangorn)
library(ggtree)

MahaMatrix <- cva.1[["Dist"]][["GroupdistMaha"]] #mahalanobis distance matrix

# Plot the Neighbor-Joining tree (unrooted by default)
nj_tree <- nj(MahaMatrix)
ggtree(nj_tree,branch.length="branch.length", cex = 0.75) + geom_tiplab(size = 2) 


# Build a rooted tree using UPGMA
upgma_tree <- upgma(MahaMatrix)
ggtree(upgma_tree,branch.length="branch.length", cex = 0.75) + geom_tiplab(size = 2) 

```

###Calculate Molecular/Morphometric Phylogenetic Tree Similarity
```{r}
#Reference Tree vs Neighbor-Joining Tree
RefvsNJ <- QuartetStatus(ref_tree, nj_tree)
RefvsNJ
SimilarityMetrics(RefvsNJ, similarity = TRUE)
VisualizeQuartets(ref_tree, nj_tree)
plot(cophylo(ref_tree, nj_tree, rotate = TRUE), link.type = "curved", link.lwd = 2, link.col = make.transparent("red", 0.5))

#Reference Tree vs UPGMA Tree
RefvsUPGMA <- QuartetStatus(ref_tree, upgma_tree)
SimilarityMetrics(RefvsUPGMA, similarity = TRUE)
VisualizeQuartets(ref_tree, upgma_tree)
plot(cophylo(ref_tree, upgma_tree, rotate = TRUE), link.type = "curved", link.lwd = 2, link.col = make.transparent("red", 0.5))

#Neighbor-Joining Tree vs UPGMA Tree
NJvsUPGMA <- QuartetStatus(nj_tree, upgma_tree)
SimilarityMetrics(NJvsUPGMA, similarity = TRUE)
plot(cophylo(nj_tree, upgma_tree, rotate = TRUE), link.type = "curved", link.lwd = 2, link.col = make.transparent("red", 0.5))
```

###Allometry
```{r}
ITD_data <- read.csv("data/ITD_covariates.csv")

lm_data <- read.csv("data/LM_Data.csv")
lm_data <- lm_data[,-1]
metadata <- lm_data[,-c(2:19)]
CVscores <- cva.1$CVscores
CV_data_full <- cbind(CVscores, metadata)


ITD_CV <- merge(CV_data_full, ITD_data, by=("catalogNumber"))

ggplot(data = ITD_CV, aes(x = ITD..cm., y = `CV 1`)) +
  geom_point(aes(color = factor(Genus)), size = 2) +
  labs(title = "Allometry Regression", x = "ITD (cm)", y = "CV1", color = "Genus") +
  geom_smooth(method = "lm")

lm_ITD <- lm(formula = `ITD..cm.`~`CV 1`, data = ITD_CV) #Linear Regression Model
summary(lm_ITD)
```


