---
title: "Bee Tree - Github"
author: "Jorge De La Cruz"
date: "2025-08-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Load Necessary Packages
```{r}
library(geomorph)
library(phytools)
library(ape)
library(ggtree)
library(vegan)
library(geiger)

packageVersion("phytools")
packageVersion("ape")

```

###Load and Prune Phylogenetic Tree
```{r}
bee_tree <- ape::read.tree("data/BEE_mat7gen_p8pmAa_fst.nwk"); phytools::read.newick("data/BEE_mat7gen_p8pmAa_fst.nwk")

?read.tree()

bee_infer <- ape::read.tree("data/BEE_mat7gen_p8pmAa_fst.nwk"); phytools::read.newick("data/BEE_mat7gen_p8pmAa_fst.nwk")


plot(bee_tree, cex = 0.1, no.margin=TRUE)

# Extract the genus from the tip labels
# Assuming the format is "genus_species~~family~~subfamily~tribe"
bee_tree$tip.label <- sub("_.*", "", bee_tree$tip.label)
bee_tree$tip.label <- sub("~.*", "", bee_tree$tip.label)
species_to_keep = c("Apis", "Bombus", "Agapostemon", "Xylocopa", "Osmia", "Megachile", "Diadasia", "Eucera", "Andrena", "Halictus", "Hesperapis", "Ceratina", "Habropoda", "Anthophora", "Panurginus", "Lasioglossum", "Calliopsis")

?write.tree()

pruned_tree <- drop.tip(bee_tree, setdiff(bee_tree$tip.label, species_to_keep)) # change "phylo_tree" to be the name of your tree


#ggtree(pruned_tree, cex = 0.1) + geom_tiplab() + ggplot2::xlim(0, 0.4)
ggtree(pruned_tree, cex = 0.5) + geom_tiplab(align = TRUE) + ggplot2::xlim(0, 0.3)

ggtree(pruned_tree, branch.length="none", cex = 1) + geom_tiplab(size = 2) 

pruned_tree

new_bee_tree <- pruned_tree

write.tree(new_bee_tree, file = "output/newick.tre")
write.nexus(new_bee_tree, file = "data/bee_tree.nex")
```
```{r}
comparePhylo(new_bee_tree, pruned_tree, plot = FALSE, force.rooted = FALSE,
             use.edge.length = FALSE, commons = TRUE,
             location = "bottomleft")
```


###Calculate Phylogenetic Distances
Extract the cophenetic (phylogenetic) distances from the phylogenetic tree.
```{r}
phylo_distances <- cophenetic.phylo(pruned_tree)

phylo_distances
```




### 2. Prepare Your Data
Ensure you have:
-A phylogenetic tree (`phylo_tree`).
-A geometric morphometric dataset (`gmm_data`) with species names corresponding to the tips of the tree.


### 4. Calculate Morphometric Distances
Calculate pairwise Procrustes distances between species.

```{r}
# Calculate mean shapes for each species
#mean_shapes <- mshape(aligned_shapes)
# Calculate Procrustes distances
#procrustes_distances <- as.matrix(dist(mean_shapes))

#mean_shapes
```


### 7. Phylogenetic Signal in Morphometric Data
Alternatively, you can test for phylogenetic signal in the 
morphometric data using methods like Blomberg’s K or Pagel’s λ.

```{r}
GM_Data <- read.csv("data/GM_Data.csv") 

row.names(GM_Data)

row.names(GM_Data) <- GM_Data$X

# Remove species column 'X'
GM_Data$X <- NULL

GM_phylo <- as.matrix(GM_Data)



#Calculate Physignal (physignal - geomorph)
bee_signal <- physignal(GM_phylo, pruned_tree)

# Estimate effect size
effect_size <- physignal.z(A = GM_phylo, phy = pruned_tree)

# Print results
summary(bee_signal)
print(effect_size)

?phylosig()

# Calculate Pagel’s λ (phylosig - phytools)
lambda_result <- phylosig(pruned_tree, GM_phylo, method = "lambda", test = TRUE)
print(lambda_result)

# Calculate Blomberg's K
K_result <- phylosig(pruned_tree, GM_phylo, method = "K", test = TRUE)
print(K_result)

# Calculate Pagel's λ (fitContinuous - geiger)
#fitContinuous(pruned_tree, GM_phylo, model="lambda")

# Calculate Pagel's λ (fitContinuous - geiger)
#fitContinuous(pruned_tree, GM_phylo, model="BM")
s
```




### Example Code
Here’s a more comprehensive example combining these steps:
```R
library(geomorph)
library(phytools)
library(ape)
# Load your data
phylo_tree <- read.tree("path_to_your_tree_file")
gmm_data <- read.csv("path_to_your_gmm_data_file")
# Perform GPA
gpa_result <- gpagen(gmm_data)
aligned_shapes <- gpa_result$coords
# Calculate mean shapes for each species
mean_shapes <- mshape(aligned_shapes)
# Calculate Procrustes distances
procrustes_distances <- as.matrix(dist(mean_shapes))
# Calculate phylogenetic distances
phylo_distances <- cophenetic.phylo(phylo_tree)
# Perform Mantel test
mantel_result <- mantel(procrustes_distances, phylo_distances)
print(mantel_result)
# Test for phylogenetic signal (Blomberg's K)
K_result <- phylosig(phylo_tree, mean_shapes, method = "K")
print(K_result)
# Test for phylogenetic signal (Pagel's lambda)
lambda_result <- phylosig(phylo_tree, mean_shapes, method = "lambda")
print(lambda_result)
```
### Notes:
- Ensure your species names match exactly between the morphometric dataset and the phylogenetic tree.
- The `phylomorphospace` function in `phytools` can visualize the phylogenetic relationships in morphospace.
This approach allows you to test the relationship between morphometric variation and phylogenetic distances, providing insight into how evolutionary history has influenced morphological traits.
