---
title: "Bee R"
author: "Jorge De La Cruz"
date: "2025-11-24"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

############################################################################################################

##Load Packages
```{r packages, warning=FALSE, echo=FALSE, message=FALSE}
library(tidyverse)
library(htmltools)
library(rgl)
library(dplyr)
library(devtools)
library(geomorph) #NOTE: Mac users please also install XQuartz from https://www.xquartz.org/. This allows the library(rgl) to function.
library(RRPP) #used for MANOVA 
library(adegenet) # used for DAPC
library(phangorn) #Phylogenetic tree Reconstruction
library(phytools) #Read phylogenetic tree files and calculate phylogenetic signal
library(ape) #Phylogenetic tree reconstruction
library(ggplot2) #Required for ggtree extension
library(ggtree) #Display phylogenetic trees
library(TreeTools) #Dependency for Quartet
library(Quartet) #Calculate quartet similarity of phylogenetic trees
library(Morpho) #package for geometric morphometrics and canonical variate analysis (CVA)
library(car) #required to plot CVA
library(shapes) #required to plot CVA

#if "ggtree" does not install use "BiocManager"
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("ggtree")
```

##Load TPS
```{r}
#Read in the TPS data for all bee wing images

all_tps_data <- readland.tps("data/bee_wings.TPS",
                             specID = c("imageID"), readcurves = FALSE, 
                             warnmsg = TRUE,negNA = FALSE)

```

##Load csv Data
Read bee metadata to be merged with the TPS data.
```{r}
bee_data <- read_csv("data/bee_metadata.csv")

#df1 <- bee_data
#df1 <- df1 %>% separate(catalogNumber, c("UCSB", "number", "side")) #issue btwn _ vs - in catalogNumber
#bee_data_clean <- df1 %>% tidyr::unite('catalogNumber', sep = "_", c('UCSB','number', 'side')) #solving it by switching to _ from - in #the downloaded Hymenoptera data

bee_data_clean <- bee_data
```

### Generalized Procrustes Analaysis (GPA)
Run GPA on TPS data, to then generate coordinate data for each landmark on each specimen. In this code, we start with tps data and end with a CSV of coordinate data for each specimen. This can then be merged with the bee data generated above.
```{r}
all_tps_gpa <- gpagen(all_tps_data, PrinAxes = TRUE)

write.csv(all_tps_gpa$coords, "data/tpsGPA.csv", row.names = TRUE)

```

### Adjust the Landmark Data to usable format
```{r}
lmdata <- read.csv("data/tpsGPA.csv") #read landmark data in

df_transpose = t(lmdata) #transpose the landmark data
df_transpose1 <- df_transpose
df_transpose1 <- df_transpose1[-c(1),] #Remove first row
#Format of LM data is currently two rows per specimen
#with one row being X coordinate values and one row
#being Y coordinate values. Below, we split into two 
#data frames and then stitch them back together to get
#18 different variables for the 9 landmarks
#(9 X coordinates and 9 Y coordinates)
lm1 <- df_transpose1
lmX <- lm1
lmY <- lm1
lmX1 <- lmX
lmX1 <- data.frame(lmX1)
lmX2 <- setNames(cbind(rownames(lmX1), lmX1, row.names = NULL),
                 c("name", "LM1x", "LM2x", "LM3x",
                   "LM4x", "LM5x", "LM6x", "LM7x", "LM8x", "LM9x"))
lmXonly <- lmX2[str_detect(lmX2$name, "X."), ]
#repeat with Y
lmY1 <- lmY
lmY1 <- data.frame(lmY1)
lmY2 <- setNames(cbind(rownames(lmY1), lmY1, row.names = NULL),
                 c("name", "LM1y", "LM2y", "LM3y",
                   "LM4y", "LM5y", "LM6y", "LM7y", "LM8y", "LM9y"))
lmYonly <- lmY2[str_detect(lmY2$name, "Y."), ]
lmXonly1 <- lmXonly
lmYonly1 <- lmYonly
lmXonly2 <- lmXonly1 %>% 
    tidyr::separate(name,                      
                    c("X","collection", "barcode", "wing","species", 
                      "location", "wingSide"), extra='drop') %>%
    tidyr::unite('catalogNumber', c('collection', 'barcode', 'wing')) 
lmXonly3 <- lmXonly2[,c(2,6:14)]

##now: lmXonly5 has x coordinate values for all 9 landmarks, and just the catalogNumber
###repeat for Y
lmYonly2 <- lmYonly1 %>% 
    tidyr::separate(name,                      
                    c("Y","collection", "barcode", "wing","species", 
                      "location", "wingSide"), extra='drop') %>%
    tidyr::unite('catalogNumber', c('collection', 'barcode', "wing")) 
lmYonly3 <- lmYonly2[,c(2,6:14)]

#Now: unite the Y and X coordinate dataframes into just one
lm_both <- merge(lmXonly3, lmYonly3, by=("catalogNumber"))

```


### Merge Landmark and Curation Data
```{r}
lm_data <- merge(lm_both, bee_data_clean, by=c("catalogNumber"), all = FALSE)

write.csv(lm_data, file = "data/LM_Data.csv")
```


### Create subset with equal numbers of specimens
Since there are different numbers of bee wings from each genus, we randomly select the same number of bees from each genus to create a subset with equal number of wings.
```{r}
set.seed(339)  # for reproducibility so that the same subset is produced every time the code is run, if removed the code will produce a new random dataset

subset <- lm_data[!lm_data$Species %in% c("Andrena oenotherae", "Bombus vosnesenskii", "Ceratina punctagena", "Megachile rotundata"), ] # remove any extra species in dataset

subset2 <- unique(subset) # remove duplicates

ITD_data <- read.csv("data/ITD_covariates.csv")

subset3 <- merge(subset2, ITD_data, by=("catalogNumber"), all = TRUE) # merge land mark data with ITD data

subset_ITD <- subset3[!is.na(subset3$Species),]

# Include all specimens with ITD values and however many specimens without ITD values until a total subset of 10 specimens is reached
subset_LM <- subset_ITD %>% 
  group_by(Genus) %>%
  group_modify(~{
    
    # Split into non-NA and NA
    non_na <- .x %>% filter(!is.na(ITD))
    na_rows <- .x %>% filter(is.na(ITD))
    
    if (nrow(non_na) >= 10) {
      # If ≥10 non-NA rows, randomly choose 10 of them
      slice_sample(non_na, n = 10)
      
    } else {
      # Otherwise, take all non-NA and fill with NA rows
      n_needed <- 10 - nrow(non_na)
      
      bind_rows(
        non_na,
        slice_sample(na_rows, n = min(n_needed, nrow(na_rows)))
      )
    }
  }) %>%
  ungroup() %>% 
  relocate(Genus, .after = Species)

write.csv(subset_LM, file = "data/subset_LM.csv")
```

### MANOVA
Using MANOVA to test for difference between groups using the RRPP package.
```{r}
# MANOVA test between genera
manova_genus <- subset_LM #subset
manova_genus <- manova_genus[,c(2:23)] #drop catalog number and other non LM variables
manova_genus_data <- manova_genus[,c(1:18)] #drop genus variable
manova_genus_data <- as.matrix(manova_genus_data) #as matrix

fit_genus <- lm.rrpp(manova_genus_data ~ Genus, SS.type = "I", #Genus
               data = manova_genus, print.progress = FALSE) #run the linear model fit
fitm_genus <- manova.update(fit_genus, print.progress = FALSE, tol = 0) #run manova update

fit_subfam <- lm.rrpp(manova_genus_data ~ Subfamily, SS.type = "I", #Subfamily
               data = manova_genus, print.progress = FALSE) #run the linear model fit
fitm_subfam <- manova.update(fit_subfam, print.progress = FALSE, tol = 0) #run manova update

fit_fam <- lm.rrpp(manova_genus_data ~ Family, SS.type = "I", #Family
               data = manova_genus, print.progress = FALSE) #run the linear model fit
fitm_fam <- manova.update(fit_fam, print.progress = FALSE, tol = 0) #run manova update

# MANOVA Genus
summary(fitm_genus, test = "Pillai") #summarize with a manova table
capture.output(summary(fitm_genus, test = "Pillai"), file = "output/manova_pillai.txt")

# MANOVA Subfamily
summary(fitm_subfam, test = "Pillai") #summarize with a manova table
capture.output(summary(fitm_subfam, test = "Pillai"), file = "output/manova_pillai_subfam.txt")

# MANOVA Family
summary(fitm_fam, test = "Pillai") #summarize with a manova table
capture.output(summary(fitm_fam, test = "Pillai"), file = "output/manova_pillai_fam.txt")

```

### Discriminant Analysis of Principal Components (DAPC)

## Cross-validation using DAPC
Assign "unknown" bees to the correct group, and test for % accuracy. Tells number of PCs to retain
```{r}
set.seed(339)  # for reproducibility so that the same # of PCs is calculated every time the code is run, if removed the code will run a new calculation

lm_values_sub <- subset_LM[,c(2:19)]

#Replace lm_data$____ with taxonomic level
xval2 <- xvalDapc(lm_values_sub, subset_LM$Species, n.pca.max = 300, training.set = 0.9, result = "groupMean", center = TRUE, scale = FALSE, n.pca = NULL, n.rep = 30, xval.plot = TRUE)

xval2$`Number of PCs Achieving Highest Mean Success` #Which number of PCs retained has highest accuracy? Put this value for n.pca = ___ when doing dapc() 

xval2$`Mean Successful Assignment by Number of PCs of PCA` #What is accuracy at various PCs retained?

```

#DAPC with 1 Discriminant Function
Rather than a PCA, we run a DAPC here to visualize difference between groups.
```{r}
dapc_1 <- dapc(lm_values_sub,  
                grp = subset_LM$Genus, 
                n.pca = 6, #Axes retained in Principal Component Analysis (PCA)
                n.da = 10) #Axes retained in Discriminant Analysis

dapc_col <- c("red4", "darkblue", "forestgreen","darkcyan", "magenta4") #save color values for DAPC

scatter(dapc_1,1,1,bg="white", #1,1, = 1st discriminant function
        cstar=0, scree.pca=FALSE, posi.pca="topright",
        legend=TRUE, size.leg=4, posi.leg="topright", col = dapc_col) # plot w/ just one axis meaning just 1 DA

```


# DAPC with 2 Discriminant Functions
```{r}
dapc_LM <- subset_LM[,c(2:19)] # Only preserves LM values

dapc_2 <- dapc(dapc_LM, 
                grp = subset_LM$Genus, #Species, Genus, Subfamily, Family, 
                n.pca = 6, #Axes retained in Principal Component Analysis (PCA)
                n.da = 10) #Axes retained in Discriminant Analysis

scatter(dapc_2, 1, 2,  # 1st and 2nd discriminant functions
        scree.da = FALSE, posi.da = "topleft", 
        bg = "white", 
        cstar = 0, pch = 20, solid = 0.4, cex = 2.5, 
        scree.pca = FALSE, posi.pca = "topleft",
        legend = TRUE, posi.leg = "bottomright",
        col = dapc_col,
        clabel = 0.5)       # size of group labels

print.dapc(dapc_2)

summary.dapc(dapc_2)
```


### Bee Phylogeny

# Load and Prune Phylogenetic Tree
```{r}
bee_tree <- ape::read.tree("data/BEE_mat7gen_p8pmAa_fst.nwk"); phytools::read.newick("data/BEE_mat7gen_p8pmAa_fst.nwk")

# Extract the genus from the tip labels
# Assuming the format is "genus_species~~family~~subfamily~tribe"
bee_tree$tip.label <- sub("_.*", "", bee_tree$tip.label)
bee_tree$tip.label <- sub("~.*", "", bee_tree$tip.label)
species_to_keep = c("Apis", "Bombus", "Agapostemon", "Xylocopa", "Osmia", "Megachile", "Diadasia", "Eucera", "Andrena", "Halictus", "Hesperapis", "Ceratina", "Habropoda", "Anthophora", "Panurginus", "Lasioglossum", "Calliopsis")

pruned_tree <- drop.tip(bee_tree, setdiff(bee_tree$tip.label, species_to_keep)) 
ref_tree <- pruned_tree

write.tree(ref_tree, file = "data/bee_tree.nwk")
write.nexus(ref_tree, file = "data/bee_tree.nex")
```

#Calculate Phylogenetic Distances
Extract the cophenetic (phylogenetic) distances from the phylogenetic tree.
```{r}
phylo_distances <- cophenetic.phylo(ref_tree)

write.csv(phylo_distances, "data/phylo_distances.csv", row.names = TRUE)
```

### Canonical Variate Analysis (CVA)
```{r}
vari <- subset_LM[,2:19]
species <- as.factor(subset_LM$Species)
genus <- as.factor(subset_LM$Genus)
subfamily <- as.factor(subset_LM$Subfamily)
family <- as.factor(subset_LM$Family)
catalog <- as.factor(subset_LM$catalogNumber)

cva.0 <- CVA(vari, groups = species) # canonical variate analysis @ species level
cva.1 <- CVA(vari, groups = genus) # canonical variate analysis @ generic level
cva.2 <- CVA(vari, groups = subfamily) # canonical variate analysis @ subfamily level
cva.3 <- CVA(vari, groups = family) # canonical variate analysis @ family level

# species
cva.0
# calculate the typicality probabilities and classification accuracy
typprobs_species <- typprobClass(cva.0$CVscores, groups = species)
print(typprobs_species)
# Variance explained by the canonical roots
cva.0$Var

?CVA()
# genus
cva.1
# calculate the typicality probabilities and classification accuracy
typprobs_genus <- typprobClass(cva.1$CVscores, groups = genus)
print(typprobs_genus)
# Variance explained by the canonical roots
cva.1$Var

# subfamily
cva.2
# calculate the typicality probabilities and classification accuracy
typprobs_subfamily <- typprobClass(cva.2$CVscores, groups = subfamily)
print(typprobs_subfamily)
# Variance explained by the canonical roots
cva.2$Var

# family
cva.3
# calculate the typicality probabilities and classification accuracy
typprobs_family <- typprobClass(cva.3$CVscores, groups = family)
print(typprobs_family)
# Variance explained by the canonical roots
cva.3$Var
```

#Visualize Canonical Variate Analysis - Genus
```{r}
grp <- levels(genus)
cols <- seq_along(grp)
pchs <- seq_along(grp)

?plot

plot(cva.1$CVscores,
     col = as.numeric(genus),#color
     pch = as.numeric(genus), #symbol
     type = "n",
     asp = 1,
     xlab = paste("1st canonical axis", round(cva.1$Var[1,2],1), "%"),
     ylab = paste("2nd canonical axis", round(cva.1$Var[2,2],1), "%"),
     xlim = c(-30,35))

points(cva.1$CVscores,
       col = as.numeric(genus),#color
       pch = as.numeric(genus),#symbols
       cex = 0.6) #point size

for(i in seq_along(grp)) {
  
  group_scores <- cva.1$CVscores[genus == grp[i], 1:2]
  
  dataEllipse(group_scores[,1],
              group_scores[,2],
              levels = 0.95, #95% confidence ellipse
              add = TRUE,
              plot.points = FALSE,
              col = cols[i],
              lwd = 1, center.cex = 1)}

legend("topright",
       inset = c(0, 0),
       legend = grp,
       col = cols,
       pch = pchs,
       cex = 0.70,
       bty = "n")

#text(cva.1$CVscores, #Label each data point
#     labels = as.character(catalog), #catalog number
#     col = as.numeric(genus),
#     cex = 0.2)
```

#Visualize Canonical Variate Analysis - Subfamily
```{r}
grp <- levels(subfamily)
cols <- seq_along(grp)
pchs <- seq_along(grp)

plot(cva.2$CVscores,
     col = as.numeric(subfamily),#color
     pch = as.numeric(subfamily), #symbol
     type = "n",
     asp = 1,
     xlab = paste("1st canonical axis", round(cva.2$Var[1,2],1), "%"),
     ylab = paste("2nd canonical axis", round(cva.2$Var[2,2],1), "%"))

points(cva.2$CVscores,
       col = as.numeric(subfamily),#color
       pch = as.numeric(subfamily),#symbols
       cex = 0.6) #point size

for(i in seq_along(grp)) {
  
  group_scores <- cva.2$CVscores[subfamily == grp[i], 1:2]
  
  dataEllipse(group_scores[,1],
              group_scores[,2],
              levels = 0.95, #95% confidence ellipse
              add = TRUE,
              plot.points = FALSE,
              col = cols[i],
              lwd = 1, center.cex = 1)}

legend("topright",
       inset = c(0, 0),
       legend = grp,
       col = cols,
       pch = pchs,
       cex = 0.75,
       bty = "n")

#text(cva.1$CVscores, #Label each data point
#     labels = as.character(catalog), #catalog number
#     col = as.numeric(genus),
#     cex = 0.2)
```

#Visualize Canonical Variate Analysis - Family
```{r}
grp <- levels(family)
cols <- seq_along(grp)
pchs <- seq_along(grp)

plot(cva.3$CVscores,
     col = as.numeric(family),#color
     pch = as.numeric(family), #symbol
     type = "n",
     asp = 1,
     xlab = paste("1st canonical axis", round(cva.3$Var[1,2],1), "%"),
     ylab = paste("2nd canonical axis", round(cva.3$Var[2,2],1), "%"))

points(cva.3$CVscores,
       col = as.numeric(family),#color
       pch = as.numeric(family),#symbols
       cex = 0.6) #point size

for(i in seq_along(grp)) {
  
  group_scores <- cva.3$CVscores[family == grp[i], 1:2]
  
  dataEllipse(group_scores[,1],
              group_scores[,2],
              levels = 0.95, #95% confidence ellipse
              add = TRUE,
              plot.points = FALSE,
              col = cols[i],
              lwd = 1, center.cex = 1)}

legend("topright",
       inset = c(0, 0),
       legend = grp,
       col = cols,
       pch = pchs,
       cex = 0.75,
       bty = "n")

#text(cva.1$CVscores, #Label each data point
#     labels = as.character(catalog), #catalog number
#     col = as.numeric(genus),
#     cex = 0.2)
```


### Calculate Phylogenetic Signal
Calculate phylogenetic signal in the morphometric data by Blomberg’s K or Pagel’s λ.
```{r}
LMmeans <- cva.1$groupmeans

# Calculate Pagel’s λ (phylosig - phytools)
lambda_result <- phylosig(ref_tree, LMmeans, method = "lambda", test = TRUE)
print(lambda_result)

# Calculate Blomberg's K
K_result <- phylosig(ref_tree, LMmeans, method = "K", test = TRUE)
print(K_result)
```

### Phylogenetic Tree Reconstruction
Estimate phylogeny from mahalanobis distance matrix produced by CVA by Neighbor-Joining and UPGMA
```{r}
MahaMatrix <- cva.1[["Dist"]][["GroupdistMaha"]] #mahalanobis distance matrix

# Build a Neighbor-Joining tree (unrooted by default)
nj_tree <- nj(MahaMatrix)
ggtree(nj_tree,branch.length="branch.length", cex = 0.75) + geom_tiplab(size = 2) 
write.tree(nj_tree, file = "data/nj_tree.nwk")

# Build a rooted UPGMA tree
upgma_tree <- upgma(MahaMatrix)
ggtree(upgma_tree,branch.length="branch.length", cex = 0.75) + geom_tiplab(size = 2)
write.tree(upgma_tree, file = "data/upgma_tree.nwk")

```

###Calculate Phylogenetic Tree Similarity by Quartet
```{r}
#Reference Tree vs Neighbor-Joining Tree
RefvsNJ <- QuartetStatus(ref_tree, nj_tree)
SimilarityMetrics(RefvsNJ, similarity = TRUE)[,"SimilarityToReference"]
VisualizeQuartets(ref_tree, nj_tree)
plot(cophylo(ref_tree, nj_tree, rotate = TRUE), link.type = "curved", link.lwd = 2, link.col = make.transparent("red", 0.5))

#Calculate p-value
obs <- SimilarityMetrics(RefvsNJ)[,"ExplicitlyAgree"]

nperm <- 999

perm_stats <- numeric(nperm)

for(i in 1:nperm){

  # permute taxa labels of morphology tree
  perm_tree <- nj_tree
  perm_tree$tip.label <- sample(perm_tree$tip.label)

  # quartet comparison
  qs_perm <- QuartetStatus(ref_tree, perm_tree)

  # similarity statistic
  perm_stats[i] <- SimilarityMetrics(qs_perm)[,"ExplicitlyAgree"]
}

pval_nj <- (sum(perm_stats >= obs) + 1) / (nperm + 1)

pval_nj

#Reference Tree vs UPGMA Tree
RefvsUPGMA <- QuartetStatus(ref_tree, upgma_tree)
SimilarityMetrics(RefvsUPGMA, similarity = TRUE)
VisualizeQuartets(ref_tree, upgma_tree)
plot(cophylo(ref_tree, upgma_tree, rotate = TRUE), link.type = "curved", link.lwd = 2, link.col = make.transparent("red", 0.5))

#Calculate p-value
obs <- SimilarityMetrics(RefvsUPGMA)[,"ExplicitlyAgree"]

nperm <- 999

perm_stats <- numeric(nperm)

for(i in 1:nperm){

  # permute taxa labels of morphology tree
  perm_tree <- upgma_tree
  perm_tree$tip.label <- sample(perm_tree$tip.label)

  # quartet comparison
  qs_perm <- QuartetStatus(ref_tree, perm_tree)

  # similarity statistic
  perm_stats[i] <- SimilarityMetrics(qs_perm)[,"ExplicitlyAgree"]
}

pval_upgma <- (sum(perm_stats >= obs) + 1) / (nperm + 1)

pval_upgma
```

###Allometry Linear Regression
```{r}
metadata <- subset_LM[,-c(2:19)]

CVscores <- cva.1$CVscores

CV_ITD_data <- cbind(CVscores, metadata)
write.csv(CV_ITD_data, "data/CV_ITD_data.csv", row.names = FALSE)

CV_sub <- CV_ITD_data[!is.na(CV_ITD_data$ITD), ]

CV_sub <- CV_sub %>%
  group_by(Genus) %>%
  slice_sample(n = 5) %>%
  ungroup()

ggplot(data = CV_sub, aes(x = ITD, y = `CV 1`)) +
  geom_point(aes(color = factor(Genus)), size = 2) +
  labs(title = "Allometry Regression", x = "ITD (cm)", y = "CV1", color = "Genus") +
  geom_smooth(method = "lm") + theme_light()

ITD_regession <- lm(formula = `ITD`~`CV 1`, data = CV_sub) #Linear Regression Model
summary(ITD_regession)
```


